Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-08-09 latex/vanishing.tex
Ready for SIAM applied math?

\section{Characterizing D-Optimal Designs when $\modcov = 0$}\label{section:vanishing}
We will now prove Theorem \ref{thm:char} which characterizes D-optimal
designs when $\modcov = 0$. The necessary first-order condition for
D-optimality of Theorem \ref{thm:constrained} for $\modcov = 0$
become:

@@

\begin{figure}
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\optimalvsnot}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector $i$,
      ]

      \addplot [fill=green!80]  table [y=prior,   meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addplot [fill=blue!60]   table [y=optimal, meta=Label, x expr=\coordindex] {\optimalvsnot};

      \addlegendentry{$\sigma^2\lambda_i^{-1}$}
      \addlegendentry{optimal $\eta_i$s}

    \end{axis}
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\optimalvsnot}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector $i$,
      ]

      \addplot [fill=green!80]  table [y=prior,       meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addplot [fill=blue!60]   table [y=sub-optimal, meta=Label, x expr=\coordindex] {\optimalvsnot};

      \addlegendentry{$\sigma^2\lambda_i^{-1}$}
      \addlegendentry{sub-optimal $\eta_i$s}

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed $m=3$
    measurements. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $\sigma^{-2}m = 3$ in both panels. The
    D-optimal design (left) increases precision where it is
    lowest. The sub-optimal design (right) does not.}
  \label{fig:optimal vs not}
\end{figure}

Theorem \ref{thm:char} facilitates understanding of D-optimal designs
when $\modcov = 0$: Imagine each eigenvector of $\fwd \prcov \fwd^*$
corresponds to a graduated lab cylinder.
%% (Figure \ref{fig:cylinder})
Each cylinder $i$ is filled, a-priori, with green liquid of
$\lambda_i^{-1}$ volume units. We are allowed $m$ measurement, so we
have blue liquid of volume $\sigma^{-2}m$ units at our disposal. When
we seek a D-optimal design, we distribute the blue liquid by
repeatedly adding a drop to whatever cylinder currently has the lowest
level of liquid in it, as long as its index $i \leq m$. The result of
such a procedure is illustrated in Figure \ref{fig:optimal vs not}.
%% \begin{figure}%{r}{0.25\textwidth} 
%%     \centering
%%     \includegraphics[width=5cm, height=5cm]{cylinders.jpg}
%%     \caption{Three graduated lab cylinders, corresponding to three
%%       eigenvectors. Prior eigenvalues not shown.}
%%     \label{fig:cylinder}
%% \end{figure}
\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
The necessary conditions for D-optimality of Theorem \ref{thm:char}
leave some freedom in choosing $\obs$, which might allow us to avoid a
clustered design. As an example, consider an inverse problem for which
the first two eigenvalues of the pushforward prior $\fwd \prcov\fwd^*$
are $\lambda_1 = 10, \lambda_2 = 5$, with eigenvectors
$\{\ev_i\}_{i=1}^\infty$ and $\sigma^2=1$. Consider the following
$\obs, \pobs$ (see also Figure \ref{fig:clusterization}):
\begin{equation}\label{eq:two designs}
  \obs =
  \begin{pmatrix}
    \sqrt{\frac{11}{40}}\ev_1 - \sqrt{\frac{29}{40}}\ev_2 \\
    \sqrt{\frac{11}{40}}\ev_1 + \sqrt{\frac{29}{40}}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \pobs =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}
\noindent It is easy to verify that $\obs^*\obs = \pobs^*\pobs$
and that both $\obs$ and $\pobs$ are D-optimal. However, $\obs$
is a clustered design ($\meas_3 = \meas_5 = \ev_1$), while $\pobs$ is
not.

We are now ready to answer two of the questions posed in the
introduction:

\textbf{Question \ref{q:why}}: Since $\obs$ is clustered and
D-optimal, measurement clusterization poses no obstruction to
D-optimality. Furthermore, a computer implementation of Lemma
\ref{thm:char} also generates $\obs$ as a solution to the D-optimal
design problem. Furthermore, randomized numerical simulations give
rise to D-optimal designs that exhibit clusterization more than 95\% of
the time (see code in supplementary). Given our model's genericity, we
conclude that measurement clusterization may be a generic phenomenon.

\textbf{Question \ref{q:replace}}: We showed that it is possible to
substitute a D-optimal clustered design (e.g. $\obs$) for a D-optimal
non-clustered design (e.g. $\pobs$). Presenting a general algorithm
for replacing a clustered D-optimal design with a non-clustered
D-optimal design is out of the scope of the current study. We can,
however, provide some intuition, utilizing $\obs, \pobs$ as a test
case. Let $\obs_{12}$ the design composed of the first and second rows
of $\obs$ (with a similar definition for $\pobs_{12}$), and note that,
up to permutations and sign changes, $\obs_{12}$ uniquely determines
$\obs$ as a clustered design. We find $\pobs_{12}$ by requiring that
(a) measurements in $\pobs_{12}$ have unit norm, (b) $\pobs_{12}^*
\pobs_{12}$ has the same eigenvectors as $\obs^*\obs$, (c)
$\pobs^*\pobs \approx \obs^*\obs$, and (d) $\pobs^*\pobs \neq
\obs^*\obs$. We then consider a ``new'' inverse problem, for which the
pushforward prior covariance is the covariance of the posterior
$\mu_{\textup{post}}^{\data, \pobs_{12}}$, namely

$$
\left(\left(\fwd \prcov\fwd^*\right)^{-1} + \sigma^{-2}
\pobs_{12}^*\pobs_{12}\right)^{-1}.
$$
%
Finally, we complete $\pobs$ to a D-optimal design with $m=5$
measurements via the construction outlined in Lemma
\ref{lemma:free}.

Interestingly, part (c) in the process detailed above implies that
there are actually infinitely many non-clustered D-optimal designs. It
is not at all clear why the numerical implementation of Lemma
\ref{lemma:free} results in a clustered design. An answer to this
question may further help shed light on the measurement clusterization
phenomenon.

\pgfplotstableread{
  Label    prior  $o_1o_2$  $o_3$   $o_4$   $o_5$    topper
  1         0.1      0.55       1       0       1      0.001
  2         0.2      1.45       0       1       0      0.001
  3         3.5      0          0       0       0      0.001
}\clusterization

\pgfplotstableread{
  Label    prior  $o_1o_2$  $o_3o_4$   $o_5$     topper
  1         0.1    0.75     0.8        1         0.001
  2         0.2    1.25     1.2        0         0.001
  3         3.5    0        0          0         0.001
}\noclusterization

\begin{figure}
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\clusterization}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector $i$,
      ]

      \addplot [fill=green!80]  table [y=prior, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=blue!60]   table [y=$o_1o_2$, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=red!60]    table [y=$o_3$, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=black!60]  table [y=$o_4$, meta=Label, x expr=\coordindex] {\clusterization};
      \addplot [fill=orange!60] table [y=$o_5$, meta=Label, x expr=\coordindex] {\clusterization};
      %% \addplot [fill=cyan!60]   table [y=$o_6$, meta=Label, x expr=\coordindex] {\clusterization};
      %% \addplot [fill=purple!60] table [y=$o_7$, meta=Label, x expr=\coordindex] {\clusterization};

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3$}
      \addlegendentry{$o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}
      %% \addlegendentry{$o_7$}   
    \end{axis}
  \end{tikzpicture}
  \qquad
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\noclusterization}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel=precision --- prior and posterior,
        xlabel=eigenvector $i$,
      ]

      \addplot [fill=green!80]  table [y=prior, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=blue!60]   table [y=$o_1o_2$, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=red!60]    table [y=$o_3o_4$, meta=Label, x expr=\coordindex] {\noclusterization};
      \addplot [fill=black!60]  table [y=$o_5$, meta=Label, x expr=\coordindex] {\noclusterization};
      %% \addplot [fill=orange!60] table [y=$o_4$, meta=Label, x expr=\coordindex] {\noclusterization};
      %% \addplot [fill=cyan!60]   table [y=$o_5$, meta=Label, x expr=\coordindex] {\noclusterization};
      %% \addplot [fill=purple!60] table [y=$o_6$, meta=Label, x expr=\coordindex] {\noclusterization};

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs$ in \eqref{eq:two designs})
    and non-clusterization (right, $\pobs$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs$ and $\pobs$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. In the left panel, a D-optimal
    design with clusterization is illustrated: $\meas_3 = \meas_5$. In
    the right panel, a D-optimal design without clusterization is
    illustrated. The contribution to the posterior precision due to
    $\meas_1$ and $\meas_2$ is denoted $\meas_1\meas_2$. These
    contributions are plotted together because neither $\meas_1$ nor
    $\meas_2$ are not in the direction of any eigenvector. Similarly
    for $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}

>>>>

\section{D-Optimal Designs Without Model Error}\label{section:vanishing}
Our goal in this section is to prove Theorem \ref{thm:char} which
characterizes D-optimal designs when $\modcov = 0$. The necessary
first-order condition for D-optimality of Theorem
\ref{thm:constrained} for $\modcov = 0$ become:

@@

\begin{figure}
  \centering
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\optimalvsnot}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel={\LARGE precision --- prior and posterior},
        xlabel={\LARGE eigenvector},
      ]
      \addplot [fill=green!80]  table [y=prior,   meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addplot [fill=blue!60]   table [y=optimal, meta=Label, x expr=\coordindex] {\optimalvsnot};     
      \addlegendentry[scale=1.4]{$\sigma^2\lambda_i^{-1}$}
      \addlegendentry[scale=1.4]{optimal $\eta_i$s}
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}[scale=0.85]
    \begin{axis}[
        ybar stacked,
        ymin=0,
        ymax=4,
        xtick=data,
        legend style={cells={anchor=east}, legend pos=north west, legend columns=-1},
        reverse legend=false, % set to false to get correct display, but I'd like to have this true
        xticklabels from table={\optimalvsnot}{Label},
        xticklabel style={text width=2cm,align=center},
        legend plot pos=right,
        ylabel={\LARGE precision --- prior and posterior},
        xlabel={\LARGE eigenvector} ,
      ]   
      \addplot [fill=green!80]  table [y=prior,       meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addplot [fill=blue!60]   table [y=sub-optimal, meta=Label, x expr=\coordindex] {\optimalvsnot};
      \addlegendentry[scale=1.4]{$\sigma^2\lambda_i^{-1}$}
      \addlegendentry[scale=1.4]{sub-optimal $\eta_i$s}
    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed $m=3$
    measurements. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $\sigma^{-2}m = 3$ in both panels. The
    D-optimal design (left) increases precision where it is
    lowest. The sub-optimal design (right) does not.}
  \label{fig:optimal vs not}
\end{figure}

Theorem \ref{thm:char} facilitates understanding of D-optimal designs
when $\modcov = 0$: Imagine each eigenvector of $\fwd \prcov \fwd^*$
corresponds to a graduated lab cylinder.
%% (Figure \ref{fig:cylinder})
Each cylinder $i$ is filled, a-priori, with green liquid of
$\lambda_i^{-1}$ volume units. We are allowed $m$ measurement, so we
have blue liquid of volume $\sigma^{-2}m$ units at our disposal. When
we seek a D-optimal design, we distribute the blue liquid by
repeatedly adding a drop to whatever cylinder currently has the lowest
level of liquid in it, as long as its index $i \leq m$. The result of
such a procedure is illustrated in Figure \ref{fig:optimal vs not}.
%% \begin{figure}%{r}{0.25\textwidth} 
%%     \centering
%%     \includegraphics[width=5cm, height=5cm]{cylinders.jpg}
%%     \caption{Three graduated lab cylinders, corresponding to three
%%       eigenvectors. Prior eigenvalues not shown.}
%%     \label{fig:cylinder}
%% \end{figure}

### 2023-08-09 latex/why.tex
Ready for SIAM applied math?

\section{An answer for Question \ref{q:why}: The cause of measurement clusterization}\label{section:why}
\begin{figure}[H]
    \centering
    \includegraphics[height=0.5\textwidth]{eigenvectors.eps}
    \caption{D-optimal measurement locations ($m=5$ measurements) and
      weighted eigenvectors for inversion of the initial condition of
      the 1D heat equation. Measurement locations and weighted
      eigenvectors are plotted over the computational domain $\Omega =
      [0, 1]$ (x-axis). Measurement clusterization occurs
      approximately at $0.31$ and $0.69$. These two locations are a
      compromise between zeros of eigenvectors a D-optimal design aims
      to ignore (third and up) and staying far from a zero of the
      first and second eigenvectors. Allocating $m=5$ measurements
      into two locations results in clusterization, according to the
      pigeonhole principle.}
  \label{fig:why}
\end{figure}

According to Theorem \ref{thm:char}, D-optimal designs aim to capture
a small subset of eigenvectors of the prior covariance, specifically
the $k$ eigenvectors with the highest prior variance. Our model
naturally achieves this objective by not measuring eigenvectors $k+1$
and above, as proven in Theorem \ref{thm:char}. Translating this
understanding to spatial problems, we anticipate that a D-optimal
design would favor measurement locations where eigenvectors $k+1$ and
above are either close to zero in value or possess small eigenvalues
in the prior spectrum, for some $k > 0$. To illustrate this
preference, Figure \ref{fig:why} depicts the scenario using the 1D
heat equation with homogeneous Dirichlet boundary
conditions\footnote{Additional details provided in the supplementary
material.}. The plot showcases four eigenvectors, scaled according to
their prior standard deviations. Since eigenvectors beyond the fourth
have insignificant prior eigenvalues, we exclude them from
consideration. Notably, we observe that measurements are clustered
near the zeros of the third and fourth eigenvectors, so we conclude
that $k=2$. Clusterization arises because there are only two locations
where the third and fourth eigenvectors approach zero, while the first
and second eigenvectors exhibit significantly non-zero
values. Consequently, when we allocate four measurements to these two
locations, they naturally cluster together, aligning with the
pigeonhole principle.

>>>>

\section{Answer to Question \ref{q:why}: The cause of clusterization}\label{section:why}
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{eigenvectors.pdf}
    \caption{D-optimal measurement locations ($m=5$ measurements) and
      weighted eigenvectors for inversion of the initial condition of
      the 1D heat equation. Measurement locations and weighted
      eigenvectors are plotted over the computational domain $\Omega =
      [0, 1]$ (x-axis). Measurement clusterization occurs
      approximately at $0.31$ and $0.69$. These two locations are a
      compromise between zeros of eigenvectors a D-optimal design aims
      to ignore (third and up) and staying far from a zero of the
      first and second eigenvectors. Allocating $m=5$ measurements
      into two locations results in clusterization, according to the
      pigeonhole principle.}
  \label{fig:why}
\end{figure}

According to Theorem \ref{thm:char}, D-optimal designs aim to capture
a small subset of eigenvectors of the prior covariance, specifically
the $k$ eigenvectors with the highest prior variance. Our model
naturally achieves this objective by not measuring eigenvectors $k+1$
and above, as proven in Theorem \ref{thm:char}. Translating this
understanding to spatial problems, we anticipate that a D-optimal
design would favor measurement locations where eigenvectors $k+1$ and
above are either close to zero in value or possess small eigenvalues
in the prior spectrum, for some $k > 0$. To illustrate this
preference, Figure \ref{fig:why} depicts the scenario using the 1D
heat equation with homogeneous Dirichlet boundary conditions (details
in the supplementary material). The plot showcases four eigenvectors,
scaled according to their prior standard deviations. Since
eigenvectors beyond the fourth have insignificant prior eigenvalues,
we exclude them from consideration. Notably, we observe that
measurements are clustered near the zeros of the third and fourth
eigenvectors, so we conclude that $k=2$. Clusterization arises because
there are only two locations where the third and fourth eigenvectors
approach zero, while the first and second eigenvectors exhibit
significantly non-zero values. Consequently, when we allocate four
measurements to these two locations, they naturally cluster together,
aligning with the pigeonhole principle.
